{\boldmath Pseudoscalar top-Higgs coupling: Exploration of $\mathrm{CP}$-odd observables to resolve the sign ambiguity }

### 2016-03-10 Paper_ttbarH_Draft.tex
New version

\begin{abstract}
\vspace*{4mm} We present a collection of $\mathrm{CP}$-odd observables
for the process $\ppprocess$ that are linearly dependent on the scalar
($\kp$) and pseudoscalar ($\kpt$) top-Higgs coupling and hence
sensitive to the corresponding relative sign. The proposed observables
are based on triple product (TP) correlations that we extract
from the expression for the differential cross section in terms of the
spin vectors of the top and antitop quarks. In order to explore other
possibilities, we progressively modify these TPs, first
by combining them, and then by replacing the spin vectors by the
lepton momenta or the $t$ and $\tbar$ momenta by their visible parts.
We generate Monte Carlo data sets for several benchmark scenarios,
including the Standard Model ($\kp=1$, $\kpt=0$) and
two scenarios with mixed $\mathrm{CP}$ properties ($\kp=1$, $\kpt=\pm 1$).
Assuming an integrated luminosity that is consistent with that
envisioned for the High Luminosity Large Hadron Collider,
and taking into account
only statistical uncertainties,
we find that the most
promising observable can disentangle the
``$\mathrm{CP}$-mixed'' scenarios 
with an effective separation of $\sim 20\sigma$ {\bf (N: In previous version it said ``statistical level" because we wanted 
  to stress that we are not considering systematic uncertainties.
  KK: This was why I added the new phrase, ``and taking into account
  only statistical uncertainties.''  You can certainly
  change this back if you like,
  though.  The previous way of wording it seemed not completely fluid,
  so I was trying to find a way of rewording it.)}.  In
the case of observables that do not require the reconstruction of the $t$
and $\tbar$ momenta, the power of discrimination is up to $\sim
16\sigma$ for the same number of events. We also show that the
most promising observables can still disentangle the
$\mathrm{CP}$-mixed scenarios when the number of events
is reduced to values consistent with expectations
for the Large Hadron Collider in the near term.

@@

In Sec.~\ref{sec3.1} we defined a fourth triple product,
$\epsilon_4 = \epsilon_3-\epsilon_2$.  We have constructed
an angular distribution related to this TP as well.
Specifically, we have analyzed the
$\Delta\phi(n_t,n_{\tbar})$ distribution in the $Q$ rest frame,
taking
$H$ to define the $z$-axis.  We have studied the distributions
for various values of $\kp$ and $\kpt$ and have
found that they are not well described by Eq.~(\ref{eq27}).
Instead of resembling sinusoids that are shifted to the left or
right for different values of the parameters, the distributions
remain peaked near $\Delta\phi(n_t,n_{\tbar})=0$, but become
slightly distorted in such a way that there is a non-zero
asymmetry [see Eq.~(\ref{eq26})].  Interestingly, while there
is not much visible change in the $\epsilon_4$ angular distribution
for different
parameter values, the associated asymmetry values are
larger than the asymmetries for the other TPs [see
  Tables \ref{table1} and \ref{table2}].  Since this angular
distribution does not exhibit much visible change for
different values of $\kp$ and $\kpt$, we do not include the corresponding
plots here.

@@

The results obtained for the TPs $\epsilon_1=
\epsilon(t,\tbar,n_t,n_{\tbar})$, $\epsilon_2=
\epsilon(Q,\tbar,n_t,n_{\tbar})$ and $\epsilon_3
=\epsilon(Q,t,n_t,n_{\tbar})$ introduced in Sec.~\ref{sec2} are
displayed in Table~\ref{table5}.  For each TP we list the 
mean value divided by the
corresponding statistical uncertainty.
%A deviation  of the estimator of $\langle
%\epsilon \rangle$, $\bar{\epsilon}$.
We see that the three observables are capable of distinguishing the SM
case from both $\mathrm{CP}$-mixed cases. Furthermore,
the two $\mathrm{CP}$-mixed cases
are clearly disentangled, since the the observables are
sensitive to the sign of $\kpt$. The observables $\langle \epsilon_2
\rangle$ and $\langle \epsilon_3 \rangle$ appear to be slightly more
sensitive than $\langle \epsilon_1 \rangle$.  Also, the
mean value for the combination $\epsilon_4$ introduced in
Sec.~\ref{sec3.1} gives slightly smaller values than those listed in
Table~\ref{table5} for $\langle \epsilon_2
\rangle$ and $\langle \epsilon_3 \rangle$,
       {\bf (KK: is my explicit reference to $\epsilon_2$ and $\epsilon_3$ correct?  I think so\ldots.)}
yielding $-4.32$, $1.11$ and $7.23$ for the cases
$(\kp=1,\kpt=-1,0,1)$, respectively. As with the asymmetry, the
purely $\mathrm{CP}$-even and purely $\mathrm{CP}$-odd cases cannot be
distinguished by the mean value, since it is linear in both $\kp$ and
$\kpt$ (see Eqs.~(\ref{eq20}) and (\ref{eq28})). Comparing the results
in Table~\ref{table5} with
the results presented in Sec.~\ref{sec3.1}, we can conclude that
the sensitivity to the NP contribution is smaller for the mean values
of the TPs under consideration
than for the corresponding asymmetries.

@@

The mean values of $\epsilon_8$ for the scenarios under
consideration are comparable with the values listed in Table
\ref{table6} for $\epsilon_5$.
We have also studied the associated angular distributions.
We do not include plots of these distributions, however,
because they
do not exhibit much visible variation for the different
scenarios under consideration.  (The situation is similar
to that encountered for the angular distribution
associated with $\epsilon_4$ -- see the discussion
at the end of Sec.~\ref{sec3.2}.)
{\bf (KK: have I worded this correctly?
  Does the distribution look similar even in the
  purely CP-odd case, for example?)}

\begin{table}[H]
\caption{Asymmetry for the TP $\epsilon_{8}$ for the SM case and the
  two $\mathrm{CP}$-mixed scenarios. The
  values are obtained with $10^5$ simulated events.}
\label{table8}
\begin{center}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
%\begin{tabular}{|c|r||r|c||r|c||r|c|}
\hhline{|====|}
%\hhline{|--------|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_8)$~~&$\mathcal{A}(\epsilon_8)/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
%\hhline{|--------|}
$1$ & $-1$~~~ & $0.0331$~ & $10.5$~ \\[0.6mm]
\hline
$1$ & $0$ & $0.0023$~ & $0.7$~ \\[0.6mm]
\hline
$1$ & $1$ & $-0.0403$~~~ & $\,-12.7$~~~~ \\[0.6mm]
\hhline{|====|}
%\hhline{|--------|}
\end{tabular}
\end{center} 
\end{table}
\section{{\boldmath{$\mathrm{CP}$}}-odd observables not depending on \MakeLowercase{{\boldmath $t$}} and \MakeLowercase{{\boldmath $\tbar$}} momenta}
\label{sec5}
The observables discussed in the preceding sections all involve the
momenta of the top and/or anti-top quarks and thus require the full
reconstruction of the kinematics of the individual $t$ and $\tbar$
systems in order to be measured. Although challenging due to the
presence of the two neutrinos in the final state, this can in principle be
done by applying a kinematic reconstruction method such as the
neutrino weighting technique~\cite{atlasconf,atlascharge}. Another
possibility is to define observables that do not depend on the $t$
and $\tbar$ momenta but instead make use of the momenta of the
$b$ and $\bbar$ quarks to which the $t$ and $\tbar$ decay. In
order to construct such observables we will take as our starting
point the most
sensitive observables studied in Secs.~\ref{sec3} and \ref{sec4},
namely those associated with the TPs $\epsilon_4$ and $\epsilon_8$,
respectively.

>>>>

\begin{abstract}
\vspace*{4mm} We present a collection of $\mathrm{CP}$-odd observables
for the process $\ppprocess$ that are linearly dependent on the scalar
($\kp$) and pseudoscalar ($\kpt$) top-Higgs coupling and hence
sensitive to the corresponding relative sign. The proposed observables
are based on triple product (TP) correlations that we extract
from the expression for the differential cross section in terms of the
spin vectors of the top and antitop quarks. In order to explore other
possibilities, we progressively modify these TPs, first
by combining them, and then by replacing the spin vectors by the
lepton momenta or the $t$ and $\tbar$ momenta by their visible parts.
We generate Monte Carlo data sets for several benchmark scenarios,
including the Standard Model ($\kp=1$, $\kpt=0$) and
two scenarios with mixed $\mathrm{CP}$ properties ($\kp=1$, $\kpt=\pm 1$).
Assuming an integrated luminosity that is consistent with that
envisioned for the High Luminosity Large Hadron Collider,
and taking into account
only statistical uncertainties,
we find that the most
promising observable can disentangle the
``$\mathrm{CP}$-mixed'' scenarios 
with an effective separation of $\sim 20\sigma$. In
the case of observables that do not require the reconstruction of the $t$
and $\tbar$ momenta, the power of discrimination is up to $\sim
16\sigma$ for the same number of events. We also show that the
most promising observables can still disentangle the
$\mathrm{CP}$-mixed scenarios when the number of events
is reduced to values consistent with expectations
for the Large Hadron Collider in the near term.

@@

In Sec.~\ref{sec3.1} we defined a fourth triple product,
$\epsilon_4 = \epsilon_3-\epsilon_2$.  {\bf We have constructed
an angular distribution related to this TP as well.
Specifically, we have analyzed the
$\Delta\phi(n_t,n_{\tbar})$ distribution in the $Q$ rest frame,
taking
$H$ to define the $z$-axis.  We have studied the distributions
for various values of $\kp$ and $\kpt$ and have
found that they are not well described by Eq.~(\ref{eq27}).
Instead of resembling sinusoids that are shifted to the left or
right for different values of the parameters, the distributions
become distorted in such a way that there is a non-zero
asymmetry [see Eq.~(\ref{eq26})].  Moreover, the associated asymmetry values are
larger than the asymmetries for the other TPs [see 
Tables \ref{table1} and \ref{table2}].}

@@

The results obtained for the TPs $\epsilon_1=
\epsilon(t,\tbar,n_t,n_{\tbar})$, $\epsilon_2=
\epsilon(Q,\tbar,n_t,n_{\tbar})$ and $\epsilon_3
=\epsilon(Q,t,n_t,n_{\tbar})$ introduced in Sec.~\ref{sec2} are
displayed in Table~\ref{table5}.  For each TP we list the 
mean value divided by the
corresponding statistical uncertainty.
%A deviation  of the estimator of $\langle
%\epsilon \rangle$, $\bar{\epsilon}$.
We see that the three observables are capable of distinguishing the SM
case from both $\mathrm{CP}$-mixed cases. Furthermore,
the two $\mathrm{CP}$-mixed cases
are clearly disentangled, since the the observables are
sensitive to the sign of $\kpt$. The observables $\langle \epsilon_2
\rangle$ and $\langle \epsilon_3 \rangle$ appear to be slightly more
sensitive than $\langle \epsilon_1 \rangle$.  Also, the
mean value for the combination $\epsilon_4$ introduced in
Sec.~\ref{sec3.1} {\bf is slightly less sensitive than $\langle \epsilon_1\rangle$,
 $\langle \epsilon_2\rangle$ and $\langle \epsilon_3\rangle$},
       {\bf (KK: is my explicit reference to $\epsilon_2$ and $\epsilon_3$ correct?  I think so\ldots.)}
{\bf with values $-4.32$, $1.11$ and $7.23$ for the cases
$(\kp=1,\kpt=-1,0,1)$, respectively}. As with the asymmetry, the
purely $\mathrm{CP}$-even and purely $\mathrm{CP}$-odd cases cannot be
distinguished by the mean value, since it is linear in both $\kp$ and
$\kpt$ (see Eqs.~(\ref{eq20}) and (\ref{eq28})). Comparing the results
in Table~\ref{table5} with
the results presented in Sec.~\ref{sec3.1}, we can conclude that
the sensitivity to the NP contribution is smaller for the mean values
of the TPs under consideration
than for the corresponding asymmetries.

@@

The mean values of $\epsilon_8$ for the scenarios under
consideration are comparable with the values listed in Table
\ref{table6} for $\epsilon_5$.
{\bf We have also studied the associated angular distributions. Specifically, 
we have analized the $\Delta\phi(\ell^+,\ell^-)$ distribution in the $t\tbar H$ rest frame,
taking $H$ to define the $z$-axis. The distributions obtained for different values of
$\kp$ and $\kpt$ are not well described by Eq.~(\ref{eq27}) (the situation is similar
to that encountered for the angular distribution
associated with $\epsilon_4$ -- see the discussion
at the end of Sec.~\ref{sec3.2}.).  For different values of the parameters,
the distributions become slightly distorted giving rise to a non-zero asymmetry 
[see Eq. (29)].}
\begin{table}[H]
\caption{Asymmetry for the TP $\epsilon_{8}$ for the SM case and the
  two $\mathrm{CP}$-mixed scenarios. The
  values are obtained with $10^5$ simulated events.}
\label{table8}
\begin{center}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
%\begin{tabular}{|c|r||r|c||r|c||r|c|}
\hhline{|====|}
%\hhline{|--------|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_8)$~~&$\mathcal{A}(\epsilon_8)/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
%\hhline{|--------|}
$1$ & $-1$~~~ & $0.0331$~ & $10.5$~ \\[0.6mm]
\hline
$1$ & $0$ & $0.0023$~ & $0.7$~ \\[0.6mm]
\hline
$1$ & $1$ & $-0.0403$~~~ & $\,-12.7$~~~~ \\[0.6mm]
\hhline{|====|}
%\hhline{|--------|}
\end{tabular}
\end{center} 
\end{table}
\section{{\boldmath{$\mathrm{CP}$}}-odd observables not depending on \MakeLowercase{{\boldmath $t$}} and \MakeLowercase{{\boldmath $\tbar$}} momenta}
\label{sec5}
The observables discussed in the preceding sections all involve the
momenta of the top and/or anti-top quarks and thus require the full
reconstruction of the kinematics of the individual $t$ and $\tbar$
systems in order to be measured. Although challenging due to the
presence of the two neutrinos in the final state, this can in principle be
done by applying a kinematic reconstruction method such as the
neutrino weighting technique~\cite{atlasconf,atlascharge}. Another
possibility is to define observables that do not depend on the $t$
and $\tbar$ momenta but instead make use of the momenta of the
$b$ and $\bbar$ quarks to which the $t$ and $\tbar$ decay. In
order to construct such observables we will take as our starting
point the most
sensitive observables studied in Secs.~\ref{sec3} and \ref{sec4},
namely those associated with the TPs $\epsilon_4$ and $\epsilon_8$,
respectively.
